Arrows for Parallel Computations

### 2018-01-05 paper/content/topologySkeletons.tex
arrow -> Arrow

The parallel |pipe| skeleton is semantically equivalent to folding over a list |[arr a a]| of arrows with |>>>|, but does this in parallel, meaning that the arrows do not have to reside on the same thread/machine. We implement this skeleton using the |ArrowLoop| type class which gives us the |loop :: arr (a, b) (c, b) -> arr a c| combinator which allows us to express recursive fix-point computations in which output values are fed back as input. For example %this
\begin{code}
loop (arr (\(a, b) -> (b, a:b)))
\end{code}
which is the same as
\begin{code}
loop (arr snd &&& arr (uncurry (:)))
\end{code}
defines an arrow that takes its input |a| and converts it into an infinite stream |[a]| of it. Using |loop| to our advantage gives us a first draft of a pipe implementation (Fig.~\ref{fig:pipeSimple}) by plugging in the parallel evaluation call |evalN conf fs| inside the second argument of |&&&| and then only picking the first element of the resulting list with |arr last|.

@@

Sometimes, this |pipe| definition can be a bit inconvenient, especially if we want to pipe arrows of mixed types together, i.e. |arr a b| and |arr b c|. By wrapping these two arrows inside a bigger arrow |arr (([a], [b]), [c]) (([a], [b]), [c])| suitable for |pipe|, we can define |pipe2| as in Fig.~\ref{fig:pipe2}.
\begin{figure}[tb]
\begin{code}
pipe2 :: (ArrowLoop arr, ArrowChoice arr,
    ArrowLoopParallel arr (fut (([a], [b]), [c])) (fut (([a], [b]), [c])) conf,
    Future fut (([a], [b]), [c]) conf) =>
	conf -> arr a b -> arr b c -> arr a c
pipe2 conf f g =
	(arr return &&& arr (const [])) &&& arr (const []) >>>
    pipe conf (replicate 2 (unify f g)) >>>
    arr snd >>>
    arr head
    where
        unify :: (ArrowChoice arr) => 
			arr a b -> arr b c -> arr (([a], [b]), [c]) (([a], [b]), [c])
        unify f' g' =
			(mapArr f' *** mapArr g') *** arr (const []) >>>
			arr (\((b, c), a) -> ((a, b), c))

@@

%Another version of |>>>| is:
%\begin{code}
%f parcomp g = (f . put) >>> (get . g)
%\end{code}
%It does not launch both arrows |f| and |g| in parallel, but allows for more smooth data communication between them. Basically, it is a |Future|-lifted \emph{sequential} |>>>|, a way to compose parallel Arrows efficiently.

@@

We can rewrite this functionality easily with the use of |loop| as the definition of the node function, |arr (i, r) (o, r)|, after being transformed into an arrow, already fits quite neatly into |loop|'s signature: |arr (a, b) (c, b) -> arr a c|. In each iteration we start by rotating the intermediary input from the nodes |[fut r]| with |second (rightRotate >>> lazy)| (Fig.~\ref{fig:edenlazyrightrotate}). Similarly to the |pipe| from Section~\ref{sec:pipe} (Fig.~\ref{fig:pipeSimple}), we have to feed the intermediary input into our |lazy| (Fig.~\ref{fig:edenlazyrightrotate}) arrow here, or the evaluation would fail to terminate. The reasoning is explained by \citet{Loogen2012} as a demand problem.

Next, we zip the resulting |([i], [fut r])| to |[(i, fut r)]| with |arr (uncurry zip)|. We then feed this into our parallel Arrow |arr [(i, fut r)] [(o, fut r)]| obtained by transforming our input arrow |f :: arr (i, r) (o, r)| into |arr (i, fut r) (o, fut r)| before |repeat|ing and lifting it with |loopParEvalN|. Finally we unzip the output list |[(o, fut r)]| list into |([o], [fut r])|.

Plugging this arrow |arr ([i], [fut r]) ([o], fut r)| into the definition of |loop| from earlier gives us |arr [i] [o]|, our ring arrow (Fig.~\ref{fig:ringFinal}). To make sure this algorithm has speedup on shared-memory machines as well, we pass the result of this arrow to |postLoopParEvalN conf (repeat (arr id))|.
This combinator can, for example, be used to calculate the shortest paths in a graph using Warshall's algorithm.
%Further details on this can be found in \cite{eden_cefp}.

@@

Similar to the |ring|, we start by rotating the input (Fig.~\ref{fig:edenlazyrightrotate}), but this time not only in one direction, but in two. This means that the intermediary input from the neighbour nodes has to be stored in a tuple |([[fut a]], [[fut b]])| in the second argument (loop only allows for two arguments) of our looped arrow |arr ([[c]], ([[fut a]], [[fut b]])) ([[d]], ([[fut a]], [[fut b]]))| and our rotation arrow becomes 
\begin{code}
second ((mapArr rightRotate >>> lazy) *** (arr rightRotate >>> lazy))
\end{code}
instead of the singular rotation in the ring as we rotate |[[fut a]]| horizontally and |[[fut b]]| vertically. Then, we zip the inputs for the input arrow with 
\begin{code}
arr (uncurry3 zipWith3 lazyzip3)
\end{code}
from |([[c]], ([[fut a]], [[fut b]]))| to |[[(c, fut a, fut b)]]|, which we then evaluate in parallel.

@@

-- Function performed by each worker
mult :: Int -> ((Matrix,Matrix),[Matrix],[Matrix]) -> (Matrix,[Matrix],[Matrix])
mult size ((sm1,sm2),sm1s,sm2s) = (result,toRight,toBottom)
    where toRight = take (size-1) (sm1:sm1s)
          toBottom = take (size-1) (sm2':sm2s)
          sm2' = transpose sm2
          sms = zipWith prMMTr (sm1:sm1s) (sm2':sm2s)
          result = foldl1' matAdd sms
\end{code}
\caption{Adapted matrix multiplication in Eden using a the |torus| skeleton. |prMM_torus| is the parallel matrix multiplication. |mult| is the function performed by each worker. |prMMTr| calculates $AB^T$ and is used for the (sequential) calculation in the chunks. |splitMatrix| splits the Matrix into chunks. |matAdd| calculates $A + B$. Omitted definitions can be found in \ref{fig:torus_example_rest}. }
\label{fig:torusMatMult}
\end{figure}
If we compare the trace from a call using our arrow definition of the torus (Fig.~\ref{fig:torus_parrows_trace}) with the Eden version (Fig.~\ref{fig:torus_eden_trace}) we can see that the behaviour of the arrow version and execution times are comparable. We discuss further benchmarks on larger clusters and in a more detail in the next section.
\begin{figure}[tb]
\olcomment{more nodes!!!}
	\centering
	\includegraphics[width=0.9\textwidth]{images/torus_matrix_parrows_scale}
	\caption[Matrix Multiplication with |torus| (Parrows)]{Matrix Multiplication with |torus| (PArrows).}
	\label{fig:torus_parrows_trace}
\end{figure}

>>>>

The parallel |pipe| skeleton is semantically equivalent to folding over a list |[arr a a]| of Arrows with |>>>|, but does this in parallel, meaning that the Arrows do not have to reside on the same thread/machine. We implement this skeleton using the |ArrowLoop| type class which gives us the |loop :: arr (a, b) (c, b) -> arr a c| combinator which allows us to express recursive fix-point computations in which output values are fed back as input. For example %this
\begin{code}
loop (arr (\(a, b) -> (b, a:b)))
\end{code}
which is the same as
\begin{code}
loop (arr snd &&& arr (uncurry (:)))
\end{code}
defines an Arrow that takes its input |a| and converts it into an infinite stream |[a]| of it. Using |loop| to our advantage gives us a first draft of a pipe implementation (Fig.~\ref{fig:pipeSimple}) by plugging in the parallel evaluation call |evalN conf fs| inside the second argument of |&&&| and then only picking the first element of the resulting list with |arr last|.

@@

Sometimes, this |pipe| definition can be a bit inconvenient, especially if we want to pipe Arrows of mixed types together, i.e. |arr a b| and |arr b c|. By wrapping these two Arrows inside a bigger Arrow |arr (([a], [b]), [c]) (([a], [b]), [c])| suitable for |pipe|, we can define |pipe2| as in Fig.~\ref{fig:pipe2}.
\begin{figure}[tb]
\begin{code}
pipe2 :: (ArrowLoop arr, ArrowChoice arr,
    ArrowLoopParallel arr (fut (([a], [b]), [c])) (fut (([a], [b]), [c])) conf,
    Future fut (([a], [b]), [c]) conf) =>
	conf -> arr a b -> arr b c -> arr a c
pipe2 conf f g =
	(arr return &&& arr (const [])) &&& arr (const []) >>>
    pipe conf (replicate 2 (unify f g)) >>>
    arr snd >>>
    arr head
    where
        unify :: (ArrowChoice arr) => 
			arr a b -> arr b c -> arr (([a], [b]), [c]) (([a], [b]), [c])
        unify f' g' =
			(mapArr f' *** mapArr g') *** arr (const []) >>>
			arr (\((b, c), a) -> ((a, b), c))

@@

%Another version of |>>>| is:
%\begin{code}
%f parcomp g = (f . put) >>> (get . g)
%\end{code}
%It does not launch both Arrows |f| and |g| in parallel, but allows for more smooth data communication between them. Basically, it is a |Future|-lifted \emph{sequential} |>>>|, a way to compose parallel Arrows efficiently.

@@

We can rewrite this functionality easily with the use of |loop| as the definition of the node function, |arr (i, r) (o, r)|, after being transformed into an Arrow, already fits quite neatly into |loop|'s signature: |arr (a, b) (c, b) -> arr a c|. In each iteration we start by rotating the intermediary input from the nodes |[fut r]| with |second (rightRotate >>> lazy)| (Fig.~\ref{fig:edenlazyrightrotate}). Similarly to the |pipe| from Section~\ref{sec:pipe} (Fig.~\ref{fig:pipeSimple}), we have to feed the intermediary input into our |lazy| (Fig.~\ref{fig:edenlazyrightrotate}) Arrow here, or the evaluation would fail to terminate. The reasoning is explained by \citet{Loogen2012} as a demand problem.

Next, we zip the resulting |([i], [fut r])| to |[(i, fut r)]| with |arr (uncurry zip)|. We then feed this into our parallel Arrow |arr [(i, fut r)] [(o, fut r)]| obtained by transforming our input Arrow |f :: arr (i, r) (o, r)| into |arr (i, fut r) (o, fut r)| before |repeat|ing and lifting it with |loopParEvalN|. Finally we unzip the output list |[(o, fut r)]| list into |([o], [fut r])|.

Plugging this Arrow |arr ([i], [fut r]) ([o], fut r)| into the definition of |loop| from earlier gives us |arr [i] [o]|, our ring Arrow (Fig.~\ref{fig:ringFinal}). To make sure this algorithm has speedup on shared-memory machines as well, we pass the result of this Arrow to |postLoopParEvalN conf (repeat (arr id))|.
This combinator can, for example, be used to calculate the shortest paths in a graph using Warshall's algorithm.
%Further details on this can be found in \cite{eden_cefp}.

@@

Similar to the |ring|, we start by rotating the input (Fig.~\ref{fig:edenlazyrightrotate}), but this time not only in one direction, but in two. This means that the intermediary input from the neighbour nodes has to be stored in a tuple |([[fut a]], [[fut b]])| in the second argument (loop only allows for two arguments) of our looped Arrow |arr ([[c]], ([[fut a]], [[fut b]])) ([[d]], ([[fut a]], [[fut b]]))| and our rotation Arrow becomes 
\begin{code}
second ((mapArr rightRotate >>> lazy) *** (arr rightRotate >>> lazy))
\end{code}
instead of the singular rotation in the ring as we rotate |[[fut a]]| horizontally and |[[fut b]]| vertically. Then, we zip the inputs for the input Arrow with 
\begin{code}
arr (uncurry3 zipWith3 lazyzip3)
\end{code}
from |([[c]], ([[fut a]], [[fut b]]))| to |[[(c, fut a, fut b)]]|, which we then evaluate in parallel.

@@

-- Function performed by each worker
mult :: Int -> ((Matrix,Matrix),[Matrix],[Matrix]) -> (Matrix,[Matrix],[Matrix])
mult size ((sm1,sm2),sm1s,sm2s) = (result,toRight,toBottom)
    where toRight = take (size-1) (sm1:sm1s)
          toBottom = take (size-1) (sm2':sm2s)
          sm2' = transpose sm2
          sms = zipWith prMMTr (sm1:sm1s) (sm2':sm2s)
          result = foldl1' matAdd sms
\end{code}
\caption{Adapted matrix multiplication in Eden using a the |torus| skeleton. |prMM_torus| is the parallel matrix multiplication. |mult| is the function performed by each worker. |prMMTr| calculates $AB^T$ and is used for the (sequential) calculation in the chunks. |splitMatrix| splits the Matrix into chunks. |matAdd| calculates $A + B$. Omitted definitions can be found in \ref{fig:torus_example_rest}. }
\label{fig:torusMatMult}
\end{figure}
If we compare the trace from a call using our Arrow definition of the torus (Fig.~\ref{fig:torus_parrows_trace}) with the Eden version (Fig.~\ref{fig:torus_eden_trace}) we can see that the behaviour of the Arrow version and execution times are comparable. We discuss further benchmarks on larger clusters and in a more detail in the next section.
\begin{figure}[tb]
\olcomment{more nodes!!!}
	\centering
	\includegraphics[width=0.9\textwidth]{images/torus_matrix_parrows_scale}
	\caption[Matrix Multiplication with |torus| (PArrows)]{Matrix Multiplication with |torus| (PArrows).}
	\label{fig:torus_parrows_trace}
\end{figure}

### 2018-01-05 paper/content/utilityFunctions.tex
arrow -> Arrow

Arrow versions of Eden's |shuffle|, |unshuffle| and the definition of |takeEach| are in Figure~\ref{fig:edenshuffleetc}. Similarly, Figure~\ref{fig:edenlazyrightrotate} contains the definition of arrow versions of Eden's |lazy| and |rightRotate| utility functions. Fig.~\ref{fig:lazyzip3etc} contains Eden's definition of |lazyzip3| together with the utility functions |uncurry3| and |threetotwo|.
The full definition of |farmChunk| is in Figure~\ref{fig:farmChunk}.
Eden definition of |ring| skeleton is in Figure~\ref{fig:ringEden}. It
follows \citet{Loogen2012}.

@@

The |parEval2| skeleton is defined in Figure~\ref{fig:parEval2}. 
We start by transforming the |(a, c)| input into a two-element list |[Either a c]| by first tagging the two inputs with |Left| and |Right| and wrapping the right element in a singleton list with |return| so that we can combine them with |arr (uncurry (:))|. Next, we feed this list into a parallel arrow running on two instances of |f +++ g| as described in the paper. After the calculation is finished, we convert the resulting |[Either b d]| into |([b], [d])| with |arr partitionEithers|. The two lists in this tuple contain only one element each by construction, so we can finally just convert the tuple to |(b, d)| in the last step.
\begin{figure}[h]
\begin{code}
parEval2 :: (ArrowChoice arr,
	ArrowParallel arr (Either a c) (Either b d) conf) =>
	conf -> arr a b -> arr c d -> arr (a, c) (b, d)
parEval2 conf f g =
	arr Left *** (arr Right >>> arr return) >>>
	arr (uncurry (:)) >>>
	parEvalN conf (replicate 2 (f +++ g)) >>>
	arr partitionEithers >>>
	arr head *** arr head
\end{code}
	\caption{Definition of |parEval2|.}
	\label{fig:parEval2}
\end{figure}
Furthermore, Fig.~\ref{fig:torus_example_rest} contains the omitted definitions of |prMMTr| (which calculates $AB^T$ for two matrices $A$ and $B$), |splitMatrix| (which splits the a matrix into chunks), and lastly |matAdd|, that calculates $A + B$ for two matrices $A$ and $B$.
\begin{figure}[h]
\begin{code}
prMMTr m1 m2 = [[sum (zipWith (*) row col) | col <- m2 ] | row <- m1]

@@

\section{Syntactic Sugar} \label{syntacticSugar}
Finally, we also give the definitions for some syntactic sugar for PArrows, namely |parstar| and |parand|.
For basic arrows, we have the |***| combinator (Fig.~\ref{fig:syntacticSugarArrows}) which allows us to combine two arrows |arr a b| and |arr c d| into an arrow |arr (a, c) (b, d)| which does both computations at once. This can easily be translated into a parallel version |parstar| with the use of |parEval2|, but for this we require a backend which has an implementation that does not require any configuration (hence the |()| as the |conf| parameter% in Fig.~\ref{fig:par***}
):
% \begin{figure}[h]
\begin{code}
(|***|) :: (ArrowChoice arr, ArrowParallel arr (Either a c) (Either b d) ())) =>
	arr a b -> arr c d -> arr (a, c) (b, d)
(|***|) = parEval2 ()
\end{code}
% \caption{Definition of |parstar|, the parallel version of |***|.}
% \label{fig:par***}
% \end{figure}
% With this we can analogously to the serial |&&&|
We define the parallel |parand| % (Fig.~\ref{fig:par&&&}) 
in a similar manner to its sequential pendant |&&&| (Fig.~\ref{fig:syntacticSugarArrows}):
% \begin{figure}[h]
\begin{code}
(|&&&|) :: (ArrowChoice arr, ArrowParallel arr (Either a a) (Either b c) ()) =>
	arr a b -> arr a c -> arr a (b, c)
(|&&&|) f g = (arr $ \a -> (a, a)) >>> f |***| g
\end{code} % $ %% formatting
% \caption{Definition of |parand| - the parallel version of |&&&|.}
% \label{fig:par&&&}
% \end{figure}

>>>>

Arrow versions of Eden's |shuffle|, |unshuffle| and the definition of |takeEach| are in Figure~\ref{fig:edenshuffleetc}. Similarly, Figure~\ref{fig:edenlazyrightrotate} contains the definition of Arrow versions of Eden's |lazy| and |rightRotate| utility functions. Fig.~\ref{fig:lazyzip3etc} contains Eden's definition of |lazyzip3| together with the utility functions |uncurry3| and |threetotwo|.
The full definition of |farmChunk| is in Figure~\ref{fig:farmChunk}.
Eden definition of |ring| skeleton is in Figure~\ref{fig:ringEden}. It
follows \citet{Loogen2012}.

@@

The |parEval2| skeleton is defined in Figure~\ref{fig:parEval2}. 
We start by transforming the |(a, c)| input into a two-element list |[Either a c]| by first tagging the two inputs with |Left| and |Right| and wrapping the right element in a singleton list with |return| so that we can combine them with |arr (uncurry (:))|. Next, we feed this list into a parallel Arrow running on two instances of |f +++ g| as described in the paper. After the calculation is finished, we convert the resulting |[Either b d]| into |([b], [d])| with |arr partitionEithers|. The two lists in this tuple contain only one element each by construction, so we can finally just convert the tuple to |(b, d)| in the last step.
\begin{figure}[h]
\begin{code}
parEval2 :: (ArrowChoice arr,
	ArrowParallel arr (Either a c) (Either b d) conf) =>
	conf -> arr a b -> arr c d -> arr (a, c) (b, d)
parEval2 conf f g =
	arr Left *** (arr Right >>> arr return) >>>
	arr (uncurry (:)) >>>
	parEvalN conf (replicate 2 (f +++ g)) >>>
	arr partitionEithers >>>
	arr head *** arr head
\end{code}
	\caption{Definition of |parEval2|.}
	\label{fig:parEval2}
\end{figure}
Furthermore, Fig.~\ref{fig:torus_example_rest} contains the omitted definitions of |prMMTr| (which calculates $AB^T$ for two matrices $A$ and $B$), |splitMatrix| (which splits the a matrix into chunks), and lastly |matAdd|, that calculates $A + B$ for two matrices $A$ and $B$.
\begin{figure}[h]
\begin{code}
prMMTr m1 m2 = [[sum (zipWith (*) row col) | col <- m2 ] | row <- m1]

@@

\section{Syntactic Sugar} \label{syntacticSugar}
Finally, we also give the definitions for some syntactic sugar for PArrows, namely |parstar| and |parand|.
For basic Arrows, we have the |***| combinator (Fig.~\ref{fig:syntacticSugarArrows}) which allows us to combine two Arrows |arr a b| and |arr c d| into an Arrow |arr (a, c) (b, d)| which does both computations at once. This can easily be translated into a parallel version |parstar| with the use of |parEval2|, but for this we require a backend which has an implementation that does not require any configuration (hence the |()| as the |conf| parameter% in Fig.~\ref{fig:par***}
):
% \begin{figure}[h]
\begin{code}
(|***|) :: (ArrowChoice arr, ArrowParallel arr (Either a c) (Either b d) ())) =>
	arr a b -> arr c d -> arr (a, c) (b, d)
(|***|) = parEval2 ()
\end{code}
% \caption{Definition of |parstar|, the parallel version of |***|.}
% \label{fig:par***}
% \end{figure}
% With this we can analogously to the serial |&&&|
We define the parallel |parand| % (Fig.~\ref{fig:par&&&}) 
in a similar manner to its sequential pendant |&&&| (Fig.~\ref{fig:syntacticSugarArrows}):
% \begin{figure}[h]
\begin{code}
(|&&&|) :: (ArrowChoice arr, ArrowParallel arr (Either a a) (Either b c) ()) =>
	arr a b -> arr a c -> arr a (b, c)
(|&&&|) f g = (arr $ \a -> (a, a)) >>> f |***| g
\end{code} % $ %% formatting
% \caption{Definition of |parand| - the parallel version of |&&&|.}
% \label{fig:par&&&}
% \end{figure}
